---
title: "Clustering using the kmeans algorithm"
author: "Bealy MECH"
date: "10/28/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercice 1 : Partition and Matrix
Write the R code which produces the partition matrix

### 1st solution
```{r}
data(iris)
dim(iris)
```

```{r}
X <- iris[, 1:4] # we don't consider the last column which is "Species"
library(nnet)
# la matrice de la partition
C <- class.ind(iris$Species)
```

La matrice C renvoit 1 si la ligne est dans l'espace (same species), O sinon.

```{r}
summary(iris$Species)  # There are 3 species, normally
```

```{r}
t(X) %*% C
```

```{r}
diag(t(C) %*% C)   # return the number of each species 
```

```{r}
# the matrix of gravity centers of the quantitative variables
t((t(X) %*% C)/diag(t(C) %*% C))
```

### Remarque 
* $t(C) * C$ renvoie une matrice diagonale donnant le nombre dans chaque espece, et donc $diagonal(t(C) * C)$ renvoie un vecteur composé de la diagonale.
* La dernière matrice donne la moyenne par espece et par variable.

### 2nd solution 

```{r}
X <- iris[, 1:4]
dim(X)
```

```{r}
names(X)
```

```{r}
cluster <- iris$Species
# Compute the matrix C = (Cik) where Cik=1 if i belongs to Ck, 0 otherwise 
C <- matrix(0, 150, 3)  # matrix 0 composes of 150 rows and 3 columns
for (i in seq(1, 150)) { # from 1 to 150 increase by 1
  if(cluster[i] == "setosa") {
    C[1, 1] = 1  # first column is the data of species "setosa"
  }
  if(cluster[i] == "versicolor") {
    C[i, 2] = 1  # second column is the data of species "versicolor"
  }
  if(cluster[i] == "virginica") {
    C[i, 3] = 1  # third column is the data of species "virginica"
  }
} 
# verify the matrix C
```

Compute M which represents the matrix of gravity center.

```{r}
X <- as.matrix(X)
M <- solve(t(C)%*%C)%*%t(C)%*%X
M
```

```{r}
dX <- dist(X)
classif <- hclust(d = dX, method = "ward.D2")
plot(classif)
```

```{r}
plot(rev(classif$height), type='l', main="hauteurs du dendrogramme décroissantes", ylab="classif$height", xlab="", las=1, col="blue")

points(1:length(classif$height), rev(classif$height), pch=20, col="red")
```

```{r}
c <- (cbind(1:length(classif$height), rev(classif$height)))
```

```{r}
c2 <- diff(c[,2])
res <- diff(c2)
P <- X
RSQ <- rep(0, nrow(P))
SQTot <- sum(scale(P,scale=FALSE)^2)
for (i in 1:nrow(P)){
  Cla <- as.factor(cutree(hclust(dX,"ward.D2"),i))
  RSQ[i] <- sum(t((t(sapply(1:ncol(P),function(i) tapply(P[,i],Cla,mean)))-apply(P,2,mean))^2)*as.vector(table(Cla)))/SQTot
}
plot(RSQ, type='l', col="red",xlab="nombre de classes")
points(1:length(RSQ),RSQ,pch=20, col="blue")
abline(h=1)
```

## Exercise 2 : The Bell Number

  1. Show that the number of partition of n objects verifies:
  $$B_{n+1} = \sum_{k=0}^{n} B_{k}$$
Cette formule s'appelle également \textbf{la relation d'Aitken}.


## Exercise 4 : Clustering of the crabs (library MASS)

```{r}
library(MASS)
data(crabs)
dim(crabs)
```

```{r}
names(crabs)
```

\textbf{Represent all crabs with a color corresponding to the specy and symbol to the sex}

```{r}
plot(crabs, col = "blue")
```

```{r}
pairs(crabs, col = c("blue", "red")[crabs$sp], pch = c(20, 21)[crabs$sex])
```

```{r}
crabsquant <- crabs[, 4:8]
pairs(crabsquant, col = c("blue", "red")[crabs$sp], pch = c(20, 21)[crabs$sex])
```

```{r}
# set a color for each specy and a different symbol per sex
summary(crabs$sex)  # Male and Female
```

### Use kmeans to find 4 classes in crabsquant 

```{r}
nbclasse <- 4
KM <- kmeans(crabsquant, nbclasse)
plot(crabsquant, asp = 1, pch = 19, col = rainbow(nbclasse)[KM$cluster])
points(KM$centers, pch = 8, col = rainbow(nbclasse)[KM$cluster], cex = 2)
```

```{r}
trueClasses <- paste(crabs$sex, crabs$sp, sep = "-")
table(trueClasses, KM$cluster)
```

```{r}
pairs(crabsquant, col = KM$cluster)
```

```{r}
pairs(crabsquant, col = as.factor(trueClasses))
```

### Clustering of raw data

```{r}
TrueClasses_mat <- matrix(c(1,2,3,4), 2, 2)
TrueClasses_mat
```

```{r}
colnames(TrueClasses_mat) <- levels(crabs$sex)
rownames(TrueClasses_mat) <- levels(crabs$sp)
TrueClasses_mat
```

```{r}
TrueClasses <- diag(TrueClasses_mat[crabs$sex, crabs$sp])
TrueClasses
```

Truclass_mat associe a chaque couple (sp, sex) un numéro de classe. TrueClasses_mat[crabs$sex, crabs$sp] applique a chaque couple possible des 2 variables la valeur, seul la diagonale correspond au vrai couple.

```{r}
res <- kmeans(crabsquant, 4)
str(res)  # structure of an R project
```

```{r}
res
```

```{r}
pairs(crabsquant, col = res$cluster) # we have 4 classes, so 4 colors for clustering k-means
```

```{r}
pairs(crabsquant, col = as.factor(TrueClasses))
```

```{r}
z <- kde2d(crabsquant[,2], crabsquant[,4]) 
contour(z, col = "purple")
# placons les points dans le graphe de contour z
points(crabsquant[,c(2,4)], col = c("blue", "orange")[crabs$sp], pch = c(20,21)[crabs$sex])
# placons encore les centres des classes
points(res$center[,c(2,4)], cex = 2, pch = 25, bg = "red")
```

### Clustering for raw data

```{r}
table(Kmeans = res$cluster, TrueClasses)
```

```{r}
WSS <- function(partition) {
  sum(partition$withinss)
}
sortie = matrix(rep(0,4000), nrow = 1000, ncol = 4)
WSSvector <- rep(0,1000)
for (i in 1:1000) {
  res <- kmeans(crabsquant,4)
  sortie[i,] <- res$withinss
  WSSvector[i] <- WSS(res)
}
summary(WSSvector)
```

```{r}
hist(WSSvector, breaks = 20)
```

Tained a bad result because of size effect (latent factor):
  * the size information is present in all variables
  * this has a masking effect
Solution: Divide all variables by one of them allows us to have features which are relative.


```{r}
library(MASS)
n <- dim(crabs)[1]   # the number of rows, nb d'individus
dim(crabs)   # dimension of dataset "crabs
```

```{r}
crabsquant <- crabs[,4:8]
# faisons une simple petite transformation pour enlever l'effet de taille
crabsquant2 <- (crabsquant/crabsquant[,3])[,-3]
j=0
for (i in c(1,2,4,5)) {
  j = j + 1
  names(crabsquant2)[j] <- paste(names(crabsquant)[j],"/",names(crabsquant[3]))
}
```

```{r}
res_bis <- kmeans(crabsquant2, 4)
z <- kde2d(crabsquant2[,2], crabsquant2[,4])
contour(z, col = "purple")
# placons les points dans le graphe de contour z
points(crabsquant2[,c(2,4)], col = c("blue", "orange")[crabs$sp], pch = c(20,21)[crabs$sex])
# placons encore les centres des classes
points(res_bis$center[,c(2,4)], cex = 2, pch = 25, bg = "red")
```

### Clustering for the transformed dataset

```{r}
table(Kmeans = res_bis$cluster, TrueClasses)
```

### How many clusters ?

```{r}
WSSkcluster <- rep(0,20)
for (k in 1:20) {
  WSSmax <- Inf
for (i in 1:10) {
  res <- kmeans(crabsquant2,k)
if (WSS(res) < WSSmax) {
  partition <- res
  WSSmax <- WSS(res)
}
}
  WSSkcluster[k] <- WSS(partition)
}
plot(WSSkcluster, xlab = "Number of clusters", ylab = "WSS", main = "Evolution of WSS with the number of cluster")
lines(WSSkcluster, col = "red")
```








